Exploring the tree of numerical semigroups

### 2014-09-30 Papier/article.tex
Florent table

Here we describe a new algorithm for the exploration of the tree of numerical
semigroup $\mathcal{T}$ and achieve the computation of $n_g$ for $g\leq 67$.
The cornerstone of our method is a new combinatorial representation of
numerical semigroup that is well suited for exploration of the
tree~$\mathcal{T}$ and allow large code optimization essentially based on use of
vectorial instructions and parallelization.
The goal of the paper is twofold:
first to present our new encoding of the semigroup and the associated
algorithms, and second to present the optimization technique which allows for
those kinds of algorithms to get speedups by factors of hundreds and even
thousand. We claim that these technique are fairly general for those kinds of
algorithms. As a support for the claim, we applied it to an algorithm of
N.~Borie enumerating integer vector modulo permutation groups~\cite{Borie} and
got a speedup by a factor larger than 2000 using 8 cores !

@@

Running the \Cilk version of our optimized algorithm  we have explored the tree of numerical semigroup upto genus $67$
The computation are done on $64$ cores \REM{Ajouter descriptif de la machine + temps mis}.
The values of $n_g$ for $g\leq 67$ are : 
\begin{center}
\begin{tabular}{|r|r||r|r||r|r|}
\hline
g & $n_g$ & g & $n_g$ & g & $n_g$ \\
\hline
0 & 1 & 23 & 170963 & 46 & 14463633648\\
1 & 1 & 24 & 282828 & 47 & 23527845502\\
2 & 2 & 25 & 467224 & 48 & 38260496374\\
3 & 4 & 26 & 770832 & 49 & 62200036752\\
4 & 7 & 27 & 1270267 & 50 & 101090300128\\
5 & 12 & 28 & 2091030 & 51 & 164253200784\\
6 & 23 & 29 & 3437839 & 52 & 266815155103\\
7 & 39 & 30 & 5646773 & 53 & 433317458741\\
8 & 67 & 31 & 9266788 & 54 & 703569992121\\
9 & 118 & 32 & 15195070 & 55 & 1142140736859\\
10 & 204 & 33 & 24896206 & 56 & 1853737832107\\
11 & 343 & 34 & 40761087 & 57 & 3008140981820\\
12 & 592 & 35 & 66687201 & 58 & 4880606790010\\
13 & 1001 & 36 & 109032500 & 59 & 7917344087695\\
14 & 1693 & 37 & 178158289 & 60 & 12841603251351\\
15 & 2857 & 38 & 290939807 & 61 & 20825558002053\\
16 & 4806 & 39 & 474851445 & 62 & 33768763536686\\
17 & 8045 & 40 & 774614284 & 63 & 54749244915730\\
18 & 13467 & 41 & 1262992840 & 64 & 88754191073328\\
19 & 22464 & 42 & 2058356522 & 65 & 143863484925550\\
20 & 37396 & 43 & 3353191846 & 66 & 233166577125714\\
21 & 62194 & 44 & 5460401576 & 67 & 377866907506273\\
22 & 103246 & 45 & 8888486816 & \\
\hline
\end{tabular}
\end{center}

@@

The computation of $n_g$ for $g\leq 33$ with algorithm \texttt{breadth} is very long because all the \texttt{8GB} of memory are consumed and the system must use swap memory to finish the computation. This algorithm was not launched

On the following table we illustrate the impact of parallelization. The test are also be performed on the \tt{i5-3570K} CPU. This CPU has two physics cores that are viewed as four virtual cores by the system. Therefore we have decided to test the parallelization of our algorithm using \Cilk upto 4 threads. 
The time of the one threaded algorithm  must be compared with the \texttt{$delta$+sse} version of the previous table : it illustrate the additional cost induced by the use of \Cilk technology.

  \begin{figure}[h!]
\begin{tabular}{|c|c|c|c|c|c|}
\hline
Threads & 30 & 35 & 40 & 45 & 50 \\
\hline
1 &  0.11 & 1.26 & 14.9 & 182 & 2201 \\ 
2 &  0.06 & 0.65 & 7.50 &  92 & 1110\\
3 &  0.05 & 0.44 & 5.14 &  63 & 747 \\
4 &  0.04 & 0.34 & 4.02 &  48 & 489 \\
\hline
\end{tabular}
\end{figure}

It should be mentioned that the hyper threading of the CPU has not been deactivated : the CPU clock is higher using one thread than  two and so on.

\textbf{Acknowledgment.} We would like to thank Shalom Eliahou for his presentation of the tree of numerical semigroups and for the interesting discussions on the subject.

>>>>

Here we describe a new algorithm for the exploration of the tree of numerical
semigroup $\mathcal{T}$ and achieve the computation of $n_g$ for $g\leq 67$.
The cornerstone of our method is a new combinatorial representation of
numerical semigroup that is well suited for exploration of the
tree~$\mathcal{T}$ and allow large code optimization essentially based on use of
vectorial instructions and parallelization.
The goal of the paper is twofold:
first to present our new encoding of the semigroup and the associated
algorithms, and second to present the optimization techniques which allows for
those kinds of algorithms to get speedups by factors of hundreds and even
thousands. We claim that these technique are fairly general for those kinds of
algorithms. As a support for the claim, we applied it to an algorithm of
N.~Borie enumerating integer vector modulo permutation groups~\cite{Borie} and
got a speedup by a factor larger than 2000 using 8 cores !

@@

Running the \Cilk version of our optimized algorithm we have explored the tree
of numerical semigroup upto genus $67$. The computation were done one a shared
$64$ core AMD Opteron\texttrademark{} Processor 6276. As other heavy
computation were running on the machine, we only used $32$ cores. The
computation took $18$~days. The values of $n_g$ for $g\leq 67$ are :
\begin{center}
\begin{tabular}{|r|r||r|r||r|r|}
\hline
g & $n_g$ & g & $n_g$ & g & $n_g$ \\
\hline
0 & 1 & 23 & 170963 & 46 & 14463633648\\
1 & 1 & 24 & 282828 & 47 & 23527845502\\
2 & 2 & 25 & 467224 & 48 & 38260496374\\
3 & 4 & 26 & 770832 & 49 & 62200036752\\
4 & 7 & 27 & 1270267 & 50 & 101090300128\\
5 & 12 & 28 & 2091030 & 51 & 164253200784\\
6 & 23 & 29 & 3437839 & 52 & 266815155103\\
7 & 39 & 30 & 5646773 & 53 & 433317458741\\
8 & 67 & 31 & 9266788 & 54 & 703569992121\\
9 & 118 & 32 & 15195070 & 55 & 1142140736859\\
10 & 204 & 33 & 24896206 & 56 & 1853737832107\\
11 & 343 & 34 & 40761087 & 57 & 3008140981820\\
12 & 592 & 35 & 66687201 & 58 & 4880606790010\\
13 & 1001 & 36 & 109032500 & 59 & 7917344087695\\
14 & 1693 & 37 & 178158289 & 60 & 12841603251351\\
15 & 2857 & 38 & 290939807 & 61 & 20825558002053\\
16 & 4806 & 39 & 474851445 & 62 & 33768763536686\\
17 & 8045 & 40 & 774614284 & 63 & 54749244915730\\
18 & 13467 & 41 & 1262992840 & 64 & 88754191073328\\
19 & 22464 & 42 & 2058356522 & 65 & 143863484925550\\
20 & 37396 & 43 & 3353191846 & 66 & 233166577125714\\
21 & 62194 & 44 & 5460401576 & 67 & 377866907506273\\
22 & 103246 & 45 & 8888486816 & \\
\hline
\end{tabular}
\end{center}

@@

The computation of $n_g$ for $g\leq 33$ with algorithm \texttt{breadth} is very long because all the \texttt{8GB} of memory are consumed and the system must use swap memory to finish the computation. This algorithm was not launched.

\Jean{
On the following table we illustrate the impact of parallelization. The test are also be performed on the \tt{i5-3570K} CPU. This CPU has two physics cores that are viewed as four virtual cores by the system. Therefore we have decided to test the parallelization of our algorithm using \Cilk upto 4 threads. 
The time of the one threaded algorithm  must be compared with the \texttt{$delta$+sse} version of the previous table : it illustrate the additional cost induced by the use of \Cilk technology.

  \begin{figure}[h!]
\begin{tabular}{|c|c|c|c|c|c|}
\hline
Threads & 30 & 35 & 40 & 45 & 50 \\
\hline
1 &  0.11 & 1.26 & 14.9 & 182 & 2201 \\ 
2 &  0.06 & 0.65 & 7.50 &  92 & 1110\\
3 &  0.05 & 0.44 & 5.14 &  63 & 747 \\
4 &  0.04 & 0.34 & 4.02 &  48 & 489 \\
\hline
\end{tabular}
\end{figure}
}

On the following table we illustrate the impact of parallelization. The test
were performed on a machine holding two \tt{Intel\texttrademark{}
  Xeon\texttrademark{} X5650} CPU running at $2.67~GHz$. Each of those CPU
have 6 physical cores that are able to run 12 threads. However, due to Hyper
threading, when more than 12 thread are running, the computation engine are
shared between two threads. Therefore, the speedup should by much less when
adding more core. We can see that this is indeed happening under the
horizontal bar in the following table.
  \begin{figure}[h!]
\begin{tabular}{|c|c|c|c|c|c|}
\hline
Threads & 30 & 35 & 40 & 45 & 50 \\
\hline
1 &  0.33 & 2.50 & 26.1 & 307 & 3709 \\
2 &  0.17 & 1.25 & 13.05 & 153.5 & 1110\\
3 &  0.11 & 0.71 & 8.57 & 102.6 & 747 \\
4 &  0.10 & 0.55 & 6.45 &  77.21 & 489 \\
8 &  0.10 & 0.36 & 3.55 &  40.22 & 489 \\
12 &  0.10 & 0.36 & 2.61 & 27.34 & 489 \\\hline
16 &  0.10 & 0.37 & 2.50 & 25.96 & 489 \\
24 &  0.12 & 0.51 & 3.18 & 25.27 & 489 \\
\hline
\end{tabular}
\end{figure}

\textbf{Acknowledgment.} We would like to thank Shalom Eliahou for his
presentation of the tree of numerical semigroups and for the interesting
discussions on the subject. We also would like to thanks Nathan Cohen who
suggested the two level of indirection stack optimization.
